---
layout: page
title: Programming with R
subtitle: Plotting
minutes: 30
---

```{r, include = FALSE}
source("tools/chunk-options.R")
opts_chunk$set(fig.path = "fig/01-plotting-")
```

```{r echo=FALSE}
dat<-read.csv('data/Site-01.csv')
dat$Gender[dat$Gender=="m"]<-"M"
dat$Gender[dat$Gender=="f"]<-"F"
dat$Gender<-factor(dat$Gender)

avg_eye_aneurisms<-apply(dat[,5:8],2,mean)
```
### Plotting

The mathematician Richard Hamming once said, "The purpose of computing is insight, not numbers," and the best way to develop insight is often to visualize data.
Visualization deserves an entire lecture (or course) of its own, but we can explore a few of R's plotting features. 

Let's take a look at the average inflammation over time.
Recall that we already calculated these values above using `apply(dat, 2, mean)` and saved them in the variable `avg_eye_aneurisms`.
Plotting the values is done with the function `plot`.

```{r plot-avg-aneurisms}
plot(avg_eye_aneurisms)
```

Above, we gave the function `plot` a vector of numbers corresponding to the average aneurism count per quadrant across all patients.
`plot` created a scatter plot where the y-axis is the average aneurism counts and the x-axis is the order, or index, of the values in the vector, which in this case correspond to the 4 quadrants of the eye.

Let's have a look at two other statistics: the maximum and minimum inflammation per day.

```{r plot-max-aneurisms}
max_eye_aneurism <- apply(dat[,5:8], 2, max)
plot(max_eye_aneurism)
```

```{r plot-min-aneurism}
min_eye_aneurism<- apply(dat[,5:8], 2, min)
plot(min_eye_aneurism)
```

#### Challenge

Create a plot showing the standard deviation of the aneurims data 
for each quadrant across all patients.

#### Key Points

* Use `variable <- value` to assign a value to a variable in order to record it in memory.
* Objects are created on demand whenever a value is assigned to them.
* The function `dim` gives the dimensions of a data frame.
* Use `object[x, y]` to select a single element from a data frame.
* Use `from:to` to specify a sequence that includes the indices from `from` to `to`.
* All the indexing and slicing that works on data frames also works on vectors.
* Use `#` to add comments to programs.
* Use `mean`, `max`, `min` and `sd` to calculate simple statistics.
* Use `apply` to calculate statistics across the rows or columns of a data frame.
* Use `plot` to create simple visualizations.

#### Next Steps

Our work so far has looked at the data from the first site, we have 4 samples from other locations.
We would like to check the others the same way, but typing in the same commands repeatedly is tedious and error-prone.
Since computers don't get bored (that we know of), we should create a way to do a complete analysis with a single command, and then figure out how to repeat that step once for each file.
These operations are the subjects of the next two lessons.
