---
layout: page
title: Programming with R
subtitle: Instructor's Guide
minutes: 0
---

```{r, include = FALSE}
source("tools/chunk-options.R")
```

## Legend


## Overall

This lesson is written as an introduction to R, but its real purpose is to
introduce the single most important idea in programming: how to solve problems
by building functions, each of which can fit in a programmer's working memory.
In order to teach that, we must teach people a little about the mechanics of
manipulating data with lists and file I/O so that their functions can do things
they actually care about.
Our teaching order tries to show practical uses of every idea as soon as it is
introduced; instructors should resist the temptation to explain the "other 90%"
of the language as well.

The secondary goal of this lesson is to give them a usable mental model of how
programs run (what computer science educators call a
[notional machine](reference.html#notional-machine) so that they can debug
things when they go wrong.
In particular, they must understand how function call stacks work.

The final example asks them to build a command-line tool that works with the
Unix pipe-and-filter model.
We do this because it is a useful skill and because it helps learners see that
the software they use isn't magical.
Tools like `grep` might be more sophisticated than the programs our learners can
write at this point in their careers, but it's crucial they realize this is a
difference of scale rather than kind.

## [First Timers](00-first-timers.html)

* Introduction to R: basics, data types and structures

* Note that some of the info are also available as supplementary lessons

## [Analyzing Patient Data](01-starting-with-data.html)

* Check learners are reading files from the correct location (set working
  directory); remind them of the shell lesson

* Provide shortcut for the assignment operator (`<-`) (RStudio: Alt+- on
  Windows/Linux; Option+- on Mac)

```{r echo = FALSE}
dat <- read.csv("data/inflammation-01.csv", header = FALSE)
element <- c("o", "x", "y", "g", "e", "n")
```

```{r}
# Challenge - Slicing (subsetting data)
element[4:1]  # first 4 characters in reverse order
element[-1]  # remove first character
element[-4]  # remove fourth character
element[-1:-4]  # remove firts to fourth characters
element[c(5, 1, 6)]  # new character vector
# Challenge - Subsetting data
max(dat[5, 3:7])
```

```{r eval=FALSE}
sd_day_inflammation <- apply(dat, 2, sd)
plot(sd_day_inflammation)
```

## [Addressing Data](01-supp-addressing-data.html)

* Note that the data frame `dat` is not the same set of data as in other lessons

## [Data Types and Structure](01-supp-data-structures.html)

## [Understanding Factors](01-supp-factors.html)

## [Introduction to RStudio](01-supp-intro-rstudio.html)

## [Reading and Writing .csv](01-supp-read-write-csv.html)

## [Creating Functions](02-func-R.html)

## [Analyzing Multiple Data Sets](03-loops-R.html)

## [Loops in R](03-supp-loops-in-depth.html)

## [Making Choices](04-cond-colors-R.html)

## [Making Choices](04-cond.html)

## [Basic Defensive Programming](05-testing-R.html)

## [Best Practices for Using R and Designing Programs](06-best-practices-R.html)

## [Command-Line Programs](06-cmdline.html)

## [Dynamic Reports with knitr](07-knitr-R.html)

## [Making Packages in R](08-making-packages-R.html)

