---
layout: page
title: Programming with R
subtitle: Instructor's Guide
minutes: 0
---

```{r, include = FALSE}
source("tools/chunk-options.R")
```

## Legend


## Overall

This lesson is written as an introduction to R, but its real purpose is to
introduce the single most important idea in programming: how to solve problems
by building functions, each of which can fit in a programmer's working memory.
In order to teach that, we must teach people a little about the mechanics of
manipulating data with lists and file I/O so that their functions can do things
they actually care about.
Our teaching order tries to show practical uses of every idea as soon as it is
introduced; instructors should resist the temptation to explain the "other 90%"
of the language as well.

The secondary goal of this lesson is to give them a usable mental model of how
programs run (what computer science educators call a
[notional machine](reference.html#notional-machine) so that they can debug
things when they go wrong.
In particular, they must understand how function call stacks work.

The final example asks them to build a command-line tool that works with the
Unix pipe-and-filter model.
We do this because it is a useful skill and because it helps learners see that
the software they use isn't magical.
Tools like `grep` might be more sophisticated than the programs our learners can
write at this point in their careers, but it's crucial they realize this is a
difference of scale rather than kind.

## [First Timers](00-first-timers.html)

* Introduction to R: basics, data types and structures

* Note that some of the info are also available as supplementary lessons

## [Analyzing Patient Data](01-starting-with-data.html)

* Check learners are reading files from the correct location (set working
  directory); remind them of the shell lesson

* Provide shortcut for the assignment operator (`<-`) (RStudio: Alt+- on
  Windows/Linux; Option+- on Mac)

```{r}
dat <- read.csv("data/inflammation-01.csv", header = FALSE)
element <- c("o", "x", "y", "g", "e", "n")
# Challenge - Slicing (subsetting data)
element[4:1]  # first 4 characters in reverse order
element[-1]  # remove first character
element[-4]  # remove fourth character
element[-1:-4]  # remove firts to fourth characters
element[c(5, 1, 6)]  # new character vector
# Challenge - Subsetting data
max(dat[5, 3:7])
```

```{r, eval=FALSE}
sd_day_inflammation <- apply(dat, 2, sd)
plot(sd_day_inflammation)
```

## [Addressing Data](01-supp-addressing-data.html)

* Note that the data frame `dat` is not the same set of data as in other lessons

## [Data Types and Structure](01-supp-data-structures.html)

## [Understanding Factors](01-supp-factors.html)

## [Introduction to RStudio](01-supp-intro-rstudio.html)

## [Reading and Writing .csv](01-supp-read-write-csv.html)

```{r, include = FALSE}
# Challenge - Updating values in a factor
carSpeeds <- read.csv(file = 'data/car-speeds.csv')
# Replace 'Blue' with 'Green' in cars$Color without using the stringsAsFactors or as.is arguments
carSpeeds$Color <- ifelse(as.character(carSpeeds$Color) == 'Blue',
                         'Green',
                         as.character(carSpeeds$Color))
# Convert colors back to factors
carSpeeds$Color <- as.factor(carSpeeds$Color)
```

## [Creating Functions](02-func-R.html)

```{r}
# Challenge - Create a function
fence <- function(original, wrapper) {
  answer <- c(wrapper, original, wrapper)
  return(answer)
}
```

```{r, eval=FALSE}
# Challenge - A more advanced function
analyze <- function(filename) {
  # Plots the average, min, and max inflammation over time.
  # Input is character string of a csv file.
  dat <- read.csv(file = filename, header = FALSE)
  avg_day_inflammation <- apply(dat, 2, mean)
  plot(avg_day_inflammation)
  max_day_inflammation <- apply(dat, 2, max)
  plot(max_day_inflammation)
  min_day_inflammation <- apply(dat, 2, min)
  plot(min_day_inflammation)
}

# Challenge - rescale
rescale <- function(v) {
  # Rescales a vector, v, to lie in the range 0 to 1.
  L <- min(v)
  H <- max(v)
  result <- (v - L) / (H - L)
  return(result)
}
```

```{r}
# Challenge - A function with default argument values
rescale <- function(v, lower = 0, upper = 1) {
  # Rescales a vector, v, to lie in the range lower to upper.
  L <- min(v)
  H <- max(v)
  result <- (v - L) / (H - L) * (upper - lower) + lower
  return(result)
}
answer <- rescale(dat[, 4], lower = 2, upper = 5)
min(answer)
max(answer)
answer <- rescale(dat[, 4], lower = -5, upper = -2)
min(answer)
max(answer)
```

## [Analyzing Multiple Data Sets](03-loops-R.html)

* The transition from the previous lesson to this one might be challenging for
  a very novice audience. Do not rush through the challenges, maybe drop some.

```{r}
# Challenge - Using loops
print_N <- function(N) {
  nseq <- seq(N)
  for (num in nseq) {
    print(num)
  }
}
print_N(3)

expo <- function(base, power) {
  result <- 1
  for (i in seq(power)) {
    result <- result * base
  }
  return(result)
}
expo(2, 4)

total <- function(vec) {
  #calculates the sum of the values in a vector
  vec_sum <- 0
  for (num in vec) {
    vec_sum <- vec_sum + num
  }
  return(vec_sum)
}
ex_vec <- c(4, 8, 15, 16, 23, 42)
total(ex_vec)

# Challenge - Using loops to analyze multiple files
analyze_all <- function(pattern) {
  # Runs the function analyze for each file in the current working directory
  # that contains the given pattern.
  filenames <- list.files(pattern = pattern)
  for (f in filenames) {
    analyze(f)
  }
}
```

## [Loops in R](03-supp-loops-in-depth.html)

## [Making Choices](04-cond-colors-R.html)

## [Making Choices](04-cond.html)

## [Basic Defensive Programming](05-testing-R.html)

## [Best Practices for Using R and Designing Programs](06-best-practices-R.html)

## [Command-Line Programs](06-cmdline.html)

## [Dynamic Reports with knitr](07-knitr-R.html)

## [Making Packages in R](08-making-packages-R.html)

